---
title: "Answer Key"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Certainly, there's more than one way to skin a cat. This is one set of suggested answers.

## [Introductory Part 2](part_02-data_structures.md)

1. Which data type is the vector `c("Tagesgericht" = 1.60, "Menü 1" = 2.00, "Menü 2" = 2.50, "Aktionsteller" = 3.50)`?

    It's a named numeric vector of type `double`, i.e. floating point numbers.
    
2. Can you assign this vector to the variable `prizes_students`? How could you access the costs for “Menü 2”?

    ```{r}
prizes_students <- c("Tagesgericht" = 1.60, "Menü 1" = 2.00, "Menü 2" = 2.50, "Aktionsteller" = 3.50)
prizes_students["Menü 2"]
    ```

3. Which menus cost more than two euros?

    ```{r}
prizes_students[prizes_students > 2]
    ```

4. Your colleagues order twice the “Tagesgericht”, three times “Menü 1”, three times “Menü 2”, and once “Aktionsteller”. The boss pays. How much does he or she need to pay?

    ```{r}
sum(prizes_students * c(2, 3, 3, 1))
    ```

5. The canteen wants to increase the prizes by 50 cents. Calculate the new prizes!

    ```{r}
prizes_students + 0.5
    ```

6. The prizes of the menus for employees are `c(3.10, 3.50, 4.00, 5.00)` respectively. Create a new vector called `prizes_employees` and assign the names of `prizes_students` to the names of this vector.

    ```{r}
prizes_employees <- c(3.10, 3.50, 4.00, 5.00)
names(prizes_employees) <- names(prizes_students)
prizes_employees
    ```

7. Let's make a list containing both vectors called `prizes`. If you want, you can include the prizes for guests, which are 50 cents higher than the costs for employees. Apply the rise in prices to all groups!

    ```{r}
prizes <- list(students = prizes_students, employees = prizes_employees, guests = prizes_employees + 0.5)
prizes

lapply(prizes, function(x) x + 0.5)
    ```

## [Introductory Part 3](part_03-working_with_strings.md)

1. What is the result of `strsplit(paste0("AG", "CGT", "A", "TGCT"), "GC")`?

    It's a list containing a `character` vector of length three. One of the two breaking points is generated by pasting the four strings together. Note that the breaking points are removed.

    ```{r echo=FALSE}
strsplit(paste0("AG", "CGT", "A", "TGCT"), "GC")
    ```

2. Suppose you have the vector `c("ÁlM", "AnJ", "DaS", "DmS", "JaW", "ShP", "SuB")`. Using regular expressions, how do you identify all words that contain
    1. the lowercase ‘a’,
    
        ```{r}
x <- c("ÁlM", "AnJ", "DaS", "DmS", "JaW", "ShP", "SuB")

x[stringr::str_detect(x, "a")]
        ```
    
    2. ‘a’ or ‘A’, 
    
        ```{r}
x[stringr::str_detect(x, "[Aa]")]
        ```
    
    3. any non-Latin character, 
        
        ```{r}
x[stringr::str_detect(x, "[^A-Za-z]")]
        ```
        
    4. more than two consonants in a row,
    
        ```{r}
x[stringr::str_detect(x, "[^AEIOUaeiou]{3,}")]
        ```
        
    5. end (or start) with ‘S’? 
    
        ```{r}
x[stringr::str_detect(x, "^S")]
x[stringr::str_detect(x, "S$")]        
        ```
