---
title: "Importing and Tidying Data"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
library(tidyverse)
```

## Importing Data from Delimited Files

A common standard to represent tabulated data as the comma-separated values, typically indicated by the extension `.csv`. However, the delimiters to represent the begin of a new column can be anything such as semicolons, white spaces (e.g. tab, blank).

### Importing Data from a Single File

The data of the plate mentioned in this session's introduction was saved in a comma-delimited file called ‘plate_1.csv’. You can find it in the sub-folder ‘part_10-working_with_tables_files’. We will import this file using `readr::read_csv()`, which will try to guess the data type stored in each column.

> In the file path, `.` refers to the current working directory (`getwd()`), so that we can use relative paths.

```{r}
plate_1 <- read_csv(file = "./part_10-working_with_tables_files/plate_1.csv")
plate_1
```

The column names have been taken from the first line of the file (since they were text). However, as we know the actual names, we should use them. They can be set either using `colnames(plate_1)<-` or during import. In the latter case, we can (and should) skip the first line of the file.

```{r}
plate_1 <- read_csv(file = "./part_10-working_with_tables_files/plate_1.csv",
                    col_names = str_c("conc_", c(1:4, 0)), skip = 1)
```

Generally, it is disencouraged to rely on row names to work with complex data sets since only one variable or parameter can be referenced. To be useful, each parameter or variable should be stored in a separate column. So, let's add one column specifying the sample type and one column specifying the replicate number.

```{r}
plate_1 <- plate_1 %>% 
  add_column(sample_id = rep(c("control", "treatment_A"), each = 2),
             replicate_id = rep(str_c("replicate_", c(1, 2)), times = 2),
             # insert before first column; default after last column
             .before = 1)
plate_1
```

If you had (for some reason) rownames in your table that you want to keep, use `dplyr::rownames_to_column()`. 

### Combining Data from Multiple Files with `row_bind`

A nice feature of using a programmatic approach for data analysis is that it allows retrieving (systematically) many files from multiple locations on your harddrive (e.g. reflecting multiple experiments) and to either analyze them in identical ways one by one, or to combine them for analysis. 

Certainly, we could import each file individually.

```{r eval=FALSE}
plate_2 <- read_csv(file = "./part_10-working_with_tables_files/plate_2.csv")
plate_3 <- read_csv(file = "./part_10-working_with_tables_files/plate_3.csv")
```

However, if we have many files to analyse, this approach can become tedious. So, we use `list.files(...)` to create a list of all files that we would like to include. These can even use regular expressions for pattern matching.

```{r}
plate_files.paths <- list.files(path = "./part_10-working_with_tables_files",
                                pattern = "plate_[^1]", full.names = TRUE)

plate_files <- sapply(plate_files.paths, read_csv, simplify = FALSE, USE.NAMES = TRUE)
plate_files
```

The `plate_files` object is a `list` with two `tibble` (`dplyr`'s way of `data.frame`) objects. Apparently, two columns have been swapped in `plate_3`, which could cause problems if we just pasted them. Luckily, R pays attention for us.

```{r}
plate_files %>% bind_rows(.id = "file")
```

To be mentioned here, we also (could) keep track of the file names (or paths) by specifying an `.id` column called `file`, which will take the names of the `list`. If need be, we can then make use of metadata stored in the file name or file path.

For the moment, let's combine the data with `plate_1` and call the object `plate_data`.

```{r}
plate_data <- bind_rows(plate_1, plate_files)

# remove objects no longer needed
rm(plate_files, plate_files.paths)
```

### Combining Data with `join`

We have seen that the command to combine rows is `dpylr::bind_rows(...)` in the `tidyverse` and `rbind(...)` in base R. To combine columns, we could use `dplyr::bind_cols(...)` or `cbind(...)` respectively. However, we must be absolutely sure that the rows align!

```{r eval=FALSE}
cbind(
  # half plate with conc_1 measurments
  plate_data %>% select(sample_id, replicate_id, conc_1),
  # half plate with conc_2 measurements
  plate_data %>% select(sample_id, replicate_id, conc_2)
)
```

In the following case, this fails. (Don't care too much about how to create the randomized halfs.)

```{r}
# half plate with conc_1 measurments
half_1 <- plate_data %>%
            select(sample_id, replicate_id, conc_1)
# half plate with conc_2 measurements
half_2 <- plate_data %>% sample_n(nrow(.)) %>% 
            select(sample_id, replicate_id, conc_2)

# this will be wrong
bind_cols(half_1, half_2)
```

To properly merge `half_1` and `half_2`, we need to use the colums named `sample_id` and `replicate_id` as indices. This is what `dplyr::inner_join(...)` does.

```{r}
# this will be correct
inner_join(half_1, half_2, by = c("sample_id", "replicate_id"))
```

If you want to keep rows that are in one table, but not in another, there are also the functions `dpylr::left_join()` and `dplyr::right_join()`. (Having both options is useful when you think of piping.)

```{r}
# remove objects no longer needed
rm(half_1, half_2)
```

### Importing Data from Excel

To read data from Excel, you can use `readxl::read_excel(...)` or `XLConnect::readWorksheetFromFile(...)`. (Neither is part of the `tidyverse` and probably not installed on your machine.)

However, I would disencourage to do so routinely, since `.xls` and `.xlsx` are neither open (i.e. disclosed), nor standardized file formats. This means they can be changed by Microsoft anytime and may consequently not be (immediately) supported by these packages.

Another problem you might be facing is, that within Excel, it is tempting to put multiple tables with different data side-by-side in a single worksheet. To import such worksheets  will require a lot of polishing from your side after.

> If you plan to import from Excel, keep one set of data per sheet and export as `.csv`.

## Tidying Data

Data can be tabulated in one of two ways: A tidy and a messy one.

> If each variable forms a column, and each row represents a single observation, and each cell a single value, we refer to this as **tidy data**. 

Having your data tidied is crucial for facilitating data manipulation, modelling, and visualization!

Here are some guidelines:

1. Don't use values as column headers, all column headers should be variable names.
2. Don't put multiple variables in a single column.
3. Don't store variables in both rows and columns.
4. Avoid storing multiple types of observational units in the same table.

Typically, if the table with your data is wider than long (if it is in the ‘wide format’), it's likely to encounter messy data. For example, `plate_1` is in one (of many possible) messy formats: The concentration is spread along the header row, not in a separate column. 

A tidyer ‘wide format’ of the same data would look like that:

```{r echo=FALSE}
plate_1 %>% 
  pivot_longer(cols = -one_of("replicate_id", "sample_id"), 
               names_to  = "concentration", values_to = "intensity") %>% 
  pivot_wider(names_from = sample_id, values_from = intensity) %>% 
  mutate(replicate_id  = str_c("<b>", replicate_id, "</b>"),
         concentration = str_c("<b>", concentration, "</b>"),) %>% 
  head(7) %>% 
  knitr::kable(format = "markdown", align = "c")
```

_(table abridged)_

The longer the table, the tidyer the data. Here is the ‘long format’ of `plate_1`:

```{r echo=FALSE}
plate_1 %>% 
  pivot_longer(cols = -one_of("replicate_id", "sample_id"), 
               names_to= "concentration", values_to = "intensity") %>% 
  mutate(sample_id     = str_c("<b>", sample_id, "</b>"),
         replicate_id  = str_c("<b>", replicate_id, "</b>"),
         concentration = str_c("<b>", concentration, "</b>"),) %>% 
  head(9) %>% 
  knitr::kable(format = "markdown", align = "c")
```

_(table abridged)_

The `tidyr` package (part of the `tidyverse`) provides functions to help you tidy messy data.

### Interconverting Wide and Long Table Formats

The degree to which you want to make your table ‘longer’ than ‘wide’ depends on the manipulation you want to perform. Typically, the ‘longest format’ serves as the linchpin to produce the ‘wider formats’.

For your convenience, this is an overview of the different functions that are around to interconvert wide and long table formats. (In case you encounter one of these or need help on related topics.)

| action | `tidyr` ≥ v1.0 | `tidyr` < v1.0 | `data.table` | `reshape`/`reshape2` |
|--------|------------------|------------------|--------------|----------------------|
| make wide table long (‘melting’) | `pivot_longer()` | `gather()` | `melt()`  | `melt()` |
| make long table wide (‘casting’) | `pivot_wider()`  | `spread()` | `dcast()` | `acast()`, `dcast()` |

As many people don't find the other function names intuitive, we will use `tidyr::pivot_longer(...)` and `tidyr::pivot_wider(...)`. 

```{r}
plate_1 %>% 
  # make wide table long
  pivot_longer(cols = conc_1:conc_0,        # selection of columns
               names_to  = "concentration", # name that describes the parameter in the selected headers
               values_to = "intensity"      # name that describes the value in the selected cells
  ) %>% 
  # print only top rows
  head(3)
```

Instead of selecting the columns `conc_1:conc_0`, we could have specified the columns which _not_ to gather; these are somtimes called the ‘identifying (id) columns’. The following commands evaluate to the same result.

```{r eval=FALSE}
plate_1 %>% 
  # select explicitly negative
  pivot_longer(cols = -c("sample_id", "replicate_id"))

plate_1 %>% 
  # select by column index
  pivot_longer(cols = 3:7)

plate_1 %>% 
  # select by name range
  pivot_longer(cols = conc_1:conc_0)

plate_1 %>% 
  # select explicitly positive
  pivot_longer(cols = c("conc_1", "conc_2", "conc_3", "conc_4", "conc_0"))

plate_1 %>% 
  # select using pattern
  pivot_longer(cols = starts_with("conc"))
```

We discuss selecting columns by name later.

For now, let's next widen/spread/cast the long/gathered/molten representation of `plate_1`! Maybe, we would like to see the measured intensities for `control` and `treatment_A` side-by-side given each concentration and the replicate.

```{r}
plate_1 %>% 
  # make wide table long
  pivot_longer(conc_1:conc_0, names_to = "concentration", values_to = "intensity") %>% 
  # make long table wide
  pivot_wider(names_from  = "sample_id", # column to use content as new column headers
              values_from = "intensity"  # column to use content to populate the new cells with
             ) %>% 
  # print only top rows
  head(3) 
```

The format of `plate_1` from above can be useful if we wanted to subtract the intensities measured in the control from the ones in the treated samples. Here, `dplyr::mutate(...)` creates a new column or modifies an exisiting column using the specified operations.

```{r}
plate_1 %>% 
  # make wide table long
  pivot_longer(conc_1:conc_0, names_to = "concentration", values_to = "intensity") %>% 
  # make long table wide
  pivot_wider(names_from = "sample_id", values_from = "intensity") -> plate_1.wide

plate_1.wide %>% 
  # make new column with corrected treatment_A values
  mutate(tratment_A.corrected = treatment_A - control) %>% head(3)

plate_1.wide %>% 
  # operate in place; not always advisable
  mutate(treatment_A = treatment_A - control) %>% head(3)
```

You can check out more examples in the [`tidyr` vignette on pivoting](https://tidyr.tidyverse.org/dev/articles/pivot.html).

### Uniting and Splitting Columns

The `tidyr::unite(...)` function takes multiple column names and pastes the column contents together. 

```{r}
plate_1 %>% 
  unite(col = "experiment",      # column to create
        sample_id, replicate_id, # columns to paste, can be unquoted
        sep = "/"                # separator, "_" is the default
  ) %>% 
  # print only top rows
  head(3)
```

The `tidyr::separate(...)` function does the reverse. By default, any non-alphanumeric character will be used to split the column. Note that missing pieces will be replaced with `NA`.

```{r}
plate_1 %>% 
  separate(col  = sample_id,                 # column to split, can be unquoted
           into = c("treatment", "specimen") # names of the new columns
  ) %>% 
  # print only top rows
  head(3)

plate_1 %>% 
  unite("experiment", sample_id, replicate_id, sep = "/") %>% 
  separate(experiment, into = str_c("piece_", 1:5)) %>% 
  # print only top rows
  head(3)
```

## Hands-On Excercise

The data for another plate has been saved in a file called ‘plates.RData’. When you open this file, there should be one additional object called `plate_4` and a named vector called `dose` in the ‘Global Environment’.

```{r}
load(file = "./part_10-working_with_tables_files/plates.RData")
```

1. For illustrative purposes merge `plate_4` with `plate_1` by `sample_id` only. Use `inner_join`, `left_join`, and `right_join`. Don't save the results.

```{r include=FALSE}
plate_4 %>% 
  separate(exp_id, into = c("sample_id", "replicate_id"), sep = "\\.") %>% 
  inner_join(plate_1, by = c("sample_id"))

plate_4 %>% 
  separate(exp_id, into = c("sample_id", "replicate_id"), sep = "\\.") %>% 
  left_join(plate_1, by = c("sample_id"))

plate_4 %>% 
  separate(exp_id, into = c("sample_id", "replicate_id"), sep = "\\.") %>% 
  right_join(plate_1, by = c("sample_id"))
```


2. Combine `plate_4` into `plate_data`, but be careful, the replicates (`.1` and `.2`) are actually `replicate_3` and `replicate_4`. You will probably want to change this.

```{r include=FALSE}
plate_4 %>% 
  separate(exp_id, into = c("sample_id", "replicate_id"), sep = "\\.") %>% 
  mutate(replicate_id = str_c("replicate_", as.numeric(replicate_id) + 2)) %>% 
  bind_rows(plate_data) %>% 
  arrange(sample_id, replicate_id) -> plate_data
```
