---
title: "Working with Strings"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This introduction will be on objects of type `character`. This is all text which appears between quotation marks (`""` or `''`) and is not executed by R. In informatics, such text objects are called ‘strings’.

Base R has a set of functions to combine and split strings, to find and replace text within strings etc. For the sake of consistency and its more powerful language, we will limit our discussion to the equivalent functions of the `stringr` package for more complex string operations.

`stringr` is part of the `tidyverse`.

## Properties of Strings

Note how we get the “length” of a string in R, i.e. the number of characters.

```{r}
# number of characters
nchar("apple")
# number of elements in the character vector
length("apple")

fruits <- c("apple", "banana", "lemon")
# number of elements in the character vector
length(fruits)
# number of characters in each element of the vector vs length of each element in the vector
sapply(fruits, nchar); sapply(fruits, length)
```

The `stringr` equivalent of `nchar("apple")` is `stringr::str_length("apple")`.

## Combining Strings

The `paste(...)` contracts strings (e.g. two `character` objects) into a single string (one `charachter` object).

```{r}
paste("banana", "cake")
```

If two vectors of different length are supplied, the shorter is repeated over the length of the longer one.

```{r}
paste(c("banana", "chocolate", "doughnut"), "cake")
```

Here is an illustration of the agruments `sep = ...` and `collapse = ...`.

```{r}
x <- paste(c("A", "B", "C", "D"), c(1, 2), sep = "_")
x
length(x) # a vector with four elements

x <- paste(c("A", "B", "C", "D"), c(1, 2), sep = "_", collapse = " and ")
x
length(x) # a vector with one element
```

The default separator of `paste` is a space `sep = " "`. `paste0(...)` is a shortcut for `paste(..., sep = "")`, which glues all arguments without spaces.

The `stringr` equivalent of `paste("pipet", "boy")` is `stringr::str_c("pipet", "boy")` with the default separator being no space `sep = ""`. It has an `collapse = ...` argument too.

## Splitting Strings

Strings can be split after a specific character with base R's `strsplit` function.

```{r}
strsplit(x = "one,2;three,4", split = ",")
```

In the `stringr` package, `stringr::str_split` is vectorized over either arguments, `string` or `pattern`.

```{r}
stringr::str_split(string = c("one,2;three,4", "five,six,7"), pattern = ",")
stringr::str_split(string = "one,2;three,4", pattern = c(",", ";"))
```

## Subsetting Strings

Subsetting is done by position. The first character in a string has index number 1.

To extract a substring from a string, base R has `substr`. We will not discuss it. The equivalent function in `stringr` is `stringr::str_sub`. It allows also counting from the end of a string, when negative indices are provided.

```{r}
library(stringr)

fruits <- c("apple", "banana", "lemon")

# the second letter
str_sub(fruits, start = 2, end = 2)
# the third to second to last letters
str_sub(fruits, start = 3, end = -2)
```

You can use `stringr::str_sub` to modify strings too.

```{r}
str_sub(fruits, start = 3, end = -2) <- "x"
fruits
```

To remove whitespace from the ends of a string, there is `stringr::str_trim`. 

```{r}
str_trim("   chitchat  ")
```

## Finding and Replacing Strings in Strings

In its easiest implementation, finding and replacing strings works as follows with `stringr::str_replace`.

```{r}
str_replace(string  = "One small step for a man, one giant leap for mankind!",
            pattern = "man",
            replacement = "mouse")
```

Note that `stringr::str_replace` replaces only the _first_ occurence of a the ‘pattern’. If you want to replace _all_ occurences, use `stringr::str_replace_all`.

```{r}
str_replace_all(string  = "One small step for a man, one giant leap for mankind!",
                pattern = "man",
                replacement = "mouse")
```

## Regular Expressions

In many cases we would be glad to find and replace not only words exactly matching our query, but also words that are constructed in a specific manner (that have a specific ‘pattern’). The ‘language’ in which such patterns are given is called ‘[regular expression](https://www.regular-expressions.info/tutorial.html)’ and is by far not limited to R, but applicable to many programming languages. 

_(This is a brief overview with some examples modified from the `stringr` [vignette on regular expressions](https://stringr.tidyverse.org/articles/regular-expressions.html).)_

In almost all `stringr` functions, you can use regular expressions as `pattern = ...`.

```{r}
str_replace_all(string  = "One small step for a man, one giant leap for mankind!",
                pattern = "step|leap", # match either ‘step’ or ‘leap’
                replacement = "piece of cake")
```

> In the next examples, we will use the `stringr::str_extract_all` function to return the substrings matching the regular expression. Since we provide a `character` vector with only a single element as `string = ...`, we immediatly index the returned list with `[[1]]` to save space while printing.

```{r}
str_extract_all(string  = "One small step for a man, one giant leap for mankind!", # one element only
                pattern = "step|leap")[[1]] # print it immediatly
str_extract_all(string  = c("One small step", "one giant leap"), # character vector with two elements
                pattern = "step|leap") # prints the entire list containing character vectors
```

### Placeholders

Regular expressions can have placeholders: `.` will match any character (except a newline). 

```{r}
str_extract_all(string  = "I keep pressing the space bar, but I'm still on planet Earth.",
                pattern = ".ar")[[1]]
str_extract_all(string  = "I keep pressing the space bar, but I'm still on planet Earth.",
                pattern = ".ar.")[[1]]
```

> If you need to match a fullstop, you need to use an ‘escape’ to tell the regular expression not to use the special meaning of “`.`”, but its literal meaning. The ‘escape’ sign for regular expressions is `\`, so you regex `\.`. However, `\` is also the escape sign for R. In order to _perserve_ the escape sign for the evaluation of the regular expression, you need to escape the escape sign as well. (Sounds confusing.) 

You need `\\.` to match a fullstop literally.

```{r}
str_extract_all(string  = "If you obey all the rules, you miss all the fun.",
                pattern = "fun\\.")[[1]]
```

### Quantifiers

To control how many times a pattern matches, you can use the repetition operators.

| operator | repetitions |
|----------|-------------|
| `?` | 0 or 1 |
| `+` | 1 or more |
| `*` | 0 or more |

```{r}
x <- "1899 or MDCCCXCIX is the last year of the 1890s decade."
str_extract(x, "CC?")
str_extract(x, "CC+")
str_extract_all(x, "CX*")[[1]]
```

You can also specify the number of matches using curly brackets.

| operator | repetitions |
|----------|-------------|
| `{n}` | exactly _n_ times |
| `{n,}` | _n_ or more times |
| `{n,m}` | between _n_ and _m_ |

```{r}
str_extract_all(x, "C{1,2}")[[1]]
```

By default, the repetition operators are ‘greedy’ in matching. They will match the _longest_ string possible. 

To make them match the _shortest_ string possible, put a `?` after them.

| operator | repetitions |
|----------|-------------|
| `??` | 0 or 1, prefer 0 |
| `+?` | 1 or more, match as few as possible |
| `*?` | 0 or more, match as few as possible |
| `{n,}?` | _n_ or more, match as few as possible |
| `{n,m}?` | match at least _n_, but then as few as possible |

```{r}
str_extract(x, c("C{2,3}", "C{2,3}?"))
```

### Special Characters

Escapes allow to specify chracters that are otherwise hard to type.

* `\s` matches any whitespace. This includes horizontal tabulation signs (`\t`), line feeds (`\n`), carriage returns (`\r`), as well as a variatey of space characters and other separators defined in Unicode. 

    The complement `\S` matches any non-whitespace character.

    ```{r}
str_replace_all("If   there is\tno  space in 
                
                space,\nI will leave.  ", "\\s+", " ")
    ```

Escapes can also define entire classes of characters.

* `\d` matches any digit (0, 1, 2, ..., 9). The complement, `\D`, matches any character that is not a decimal digit.

    ```{r}
str_extract_all("There are 214 moons in the solar system.", "\\d+")[[1]]
str_extract_all(x, "\\d+")[[1]]
    ```

* `\w` matches any ‘word’ character. Words consist of alphabetic characters (including diacritics etc.) and decimal numbers. 

    The complement `\W` matches any non-word character.

    ```{r}
str_extract_all("Don't shoot for the moon!", "\\w+")[[1]]
str_extract_all("Don't shoot for the moon!", "\\W+")[[1]]
    ```

* `\b` matches the transition between word and non-word characters. (Rarely needed; example below.)

You can specify your own _character classes_ using `[]`:

* `[abc]` matches a, b, or c,
* `[a-z]` matches every character between a and z (case-sensitive),
* `[^abc]` matches anything except a, b, or c.

There are a number of pre-built character classes in `stringr`:

* `[:punct:]` punctuation,
* `[:alpha:]` letters,
* `[:lower:]` lowercase letters,
* `[:upper:]` uppercase letters,
* `[:digit:]` digits,
* `[:alnum:]` letters and numbers,
* `[:cntrl:]` control chacracters,
* `[:graph:]` letters, numbers, and punctuation,
* `[:print:]` letters, numbers, punctuation, and whitespace,
* `[:space:]` space characters (same as `\s`),
* `[:blank:]` space and tab.

### Anchors

In some cases, you want to match a pattern only if it occurs at the start or the end of a line. There are

* `^` to match the start of a line, and
* `$` to match the end of the line.

```{r}
str_extract_all("papaya or banana", "[yn]a")[[1]]    # all occurences of “ya” or “na”
str_extract_all("papaya or banana", "[yn]a$")[[1]]   # matches at the end of a line only
str_extract_all("papaya or banana", "[yn]a\\b")[[1]] # matches at word boundaries only
```

To remember which one is which, [this mnemonic](https://twitter.com/emisshula/status/323863393167613953) might help you: ‘If you begin with power (`^`), you end up with money (`$`).’

## Learning Objectives

In this introduction you should have learned

* how to combine and split strings,
* how to subset strings based on character indices,
* how to replace words in strings, and
* that you can use regular expressions to formulate advanced string queries.

Quick questions.

1. What is the result of `strsplit(paste0("AG", "CGT", "A", "TGCT"), "GC")`?
2. Suppose you have the vector `c("ÁlM", "AnJ", "DaS", "DmS", "JaW", "ShP", "SuB")`. Using regular expressions, how do you identify all elements that contain
    1. the lowercase ‘a’,
    2. ‘a’ or ‘A’, 
    3. any non-Latin character, 
    4. more than two consonants in a row,
    5. end (or start) with ‘S’? 
