---
title: "Useful Statements"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Base R Statements

### Repeating Vector Elements

```{r}
rep(c(1, 2), times = 2)
rep(c(1, 2), each = 2)
```

### Making Combinations

In case you need all combinations of any sorts.

```{r}
expand.grid(Name_2 = c("a", "b", "c"), Name_1 = c(1, 2, 3, 4), c("good", "bad"))
```

### Sampling at Random

Sampling random integers.

```{r include=FALSE}
set.seed(210834)
```
```{r}
sample(1:10, size = 5, replace = TRUE)
```

Sampling values from frequency distributions.

```{r}
# from a uniform distribution
runif(5, min = 0, max = 1)
# from a normal distribution
rnorm(5, mean = 0, sd = 1)
# from a Poisson distribution
rpois(5, lambda = 2)
# from a binomial distribution
rbinom(5, size = 10, prob = .5)
```

### Working with Dates

When was the last time (and when will be the next time) in the 21st century that 1st of December is First Advent Sunday?

```{r}
year_range <- 2000:2100
year_range[which(weekdays(as.Date(paste0('0112', year_range), '%d%m%Y')) == "Sunday")]
```

## `tidyverse` Statements

```{r message=FALSE}
library(tidyverse)
```

### Matching Nearby Ranges

Floating point number representations are limited in their precision. This is why `sqrt(2)^2 == 2` is `FALSE` (within the limits of your computer's accuracy, `.Machine$double.eps`). A a safe way of comparing two vectors of floating point numbers is `dplyr::near`.

```{r}
# within machine accuracy
near(sqrt(2)^2, 2)
# within user-defined range
near(3, 2.6, tol = .5)
```

## More Pipes

See [this chapter](https://r4ds.had.co.nz/pipes.html) in G. Grolemund and H. Wickham's _R for Data Science_.