---
title: "Basic Interactions"
output: 
  github_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

After you installed R (and RStudio), it's time to get to know each other.

## The Prompt

The only meaningful way of interacting with R is by typing code into the R console. R's command prompt is a ‘>’ sign. (You don't type the ‘>’, R does.) When present, R is waiting for the next line of code. (You do hit the return key at the end of the line.)

The prompt looks like this:

```
> _
```

In the absence of the prompt, R is busy with calculations. Anything you type meanwhile, will be evaluated _after_ the current command has completed.

In case a line is syntactically incomplete after you hit the return key, R will continue the next line with a ‘+’ waiting for you to complete the command.

> If at any point you want to abort the execution of a running command, or if you cannot figure out how to correctly complete the current line of input, hit Ctrl + C (Linux or macOS command line). In RStudio, the Escape key will work on all operating systems.

If you want to recall, (modify) and re-execute a previous command, you can navigate the command history by using the up- and down-arrow keys on your keyboard. In RStudio, you can access your command history in the ‘History’ pane, which is next to the ‘Environment’ pane by default.

You quit the R session with the command `q()` or `quit()`. In RStudio, you may simply close the window.

## Code and Comments

In the following, pieces of R code (‘chunks’) are presented in grey boxes. You can copy the commands from these chunks to your R console and execute them by yourself. Beware that R is a case-sensitive language; so `A` and `a` are different symbols.

For the sake of reproducibility, some chunks are directly followed by the output that you should obtain on your machine. In this case, the lines of codes are prefixed with a ‘##’ mark.

> In R, the ‘#’ sign serves as comment character that can appear nearly anywhere in the code. Anything typed behind this sign will be invisible to R.

## A Sample Session

To break the ice, you can run the following chunk that will demonstrate one simple application of R: Fitting a linear regression model to some noisy data. 

```{r include=FALSE}
set.seed(1228)
```
```{r}
# create a “list” of values from 1 to 20
x <- 1:20
# use this list to create a “table” with two columns x and y; add
# some noise to the y observations
d <- data.frame(x = x, y = x + rnorm(x))
# make a plot
plot(d)
# fit a linear regression
r <- lm(y ~ x, data = d)
# add to plot
lines(x, r$fitted.values, col = "red")
# get the statistics of the fit
summary(r)
# clean up
rm(x, d, r)
```


## Expressions and Assignments

Code to interact with R will fall into one of two categories.

1. Expressions:

    ```{r}
1 + sin(pi/2) 
nchar("math") # number of characters in the word “math”
    ```

2. Assignments:
  
    ```{r eval=FALSE}
x = 1  # right-hand side (RHS) is assigned to left-hand side (LHS)

y <- 2 # RHS is assigned to LHS
3 -> z # LHS is assigned to RHS
    ```

When expressions are typed, R will evaluate them and print the result (unless specifically made invisible). The result of an expression is however ‘lost’ for further usage. — Assignments on the other hand do not print to the console, but store the result in the current environment or modify an existing value in the memory.

> It is not important (but good practice) to leave a space between the LHS and RHS of `=`, `->`, `<-`, and after `,`. 

### Expressions

Although expressions are not limited to mathematical expressions in the colloquial sense, R comes with a lot of  operators and functions to do anything from boolean to matrix algebra, over statistics and modelling. Thus, R can be used as a very fancy calculator. 

Some of the arithmetic operators implemented in R are shown below.

| operator | effect |
| ---------|----------|
| `+` | addition |
| `-` | subtraction |
| `*` | multiplication |
| `/` | division |
| `%/%` | integer division (i.e. without remainder) |
| `%%` | modulo (i.e. remainder only) |
| `^` or `**` | exponention |

```{r eval=FALSE}
13 %/% 4
13 %% 4

# check out the result of these ‘tricky’ cases
1/0
1/0 + 1/0
(-8)^(1/3)
```

Some of the built-in functions for other mathemtatical operations include `abs(x)`, `floor(x)`, `round(x)`, `sqrt(x)`, `exp(x)`, `log(x)`, `log10(x)`, `sin(x)`, `cos(x)`, etc. Try them out!

R follows the standard order of operations and groupings based on parentheses.

```{r eval=FALSE}
6 + 9 / 3
(6 + 9) / 3
```

### Using a Pipe

Besides grouping by parentheses, there is also another way to specify the order in which expressions are evaluated. Namely by _piping_ the different operations in the order they must be executed by R. The ‘pipe’ is designated by the `%>%` operator. 

This operator is so useful that it has its own key-binding in RStudio, which is Ctrl/Cmd + Shift + M.

We can obtain the same result as above with

```{r message=FALSE, eval=FALSE}
library(magrittr) # allow to use pipe

9 %>% divide_by(3) %>% add(6) 
6 %>% add(9) %>% divide_by(3)
```

Piping is very effective, when multiple functions would else be nested: It helps to unclutter code.

```{r eval=FALSE}
# compare ...
asin(sqrt(divide_by(sum(1, 2, 3, 4), 10)))

# ... and ...
sum(1, 2, 3, 4) %>% 
  # comments may be interspresed to give rationales
  divide_by(10) %>% 
  # functions with a single argument may omit brackets ...
  sqrt %>% 
  asin
```

### Assignments

While using R as a calculator is interesting, to do useful things, we need to assign _values to objects_. In R, the assignment operator `<-` (or `->`) points from the value to the object. The ‘=’ sign can also be used for assignments.

What's an object? It's a name that you can use to recall stored data.

Objects can be given any name (also known as the object's ‘symbol’), except that names

* cannot start with a number,
* contain some special symbols like `!`, `$`, `@`, `%`, or mathematical operators such as `+`, `-`, `*`, `/`, `^`, 
* names cannot be identical to reserved words such as `if`, `TRUE`, `NA`, `...` etc. (type `?Reserved` for a complete list).

> Beware that you might unintentionally **overwrite built-in objects** such as `mean`, `data`, `df` etc. without asking you for permission. When in doubt, check the help to see if the name is already in use. RStudio will show you a list of all currently known objects and functions that start with the characters you just typed, when you hit the ‘tab’ key. 

Although there is no restriction in length for object names, in practice, you want your names to be explicit and not too long. 

```{r}
N_Avogrado = 6.022e23 # we cannot use NA as name!

N_fmol = 2
N_abs  = N_fmol * N_Avogrado * 1e-15

# print the result
N_abs
```

You can see which objects are currently loaded in R's memory (the ‘Global Environment’) by checking the ‘Environment’ pane in RStudio or by typing `ls()` (or `objects()`) in the R console.

To remove objects from R's memory, use the `rm(...)` function.

```{r}
ls()
rm(N_abs, N_Avogrado)
ls()
```

### Functions

If a sequence of operations has to be repeatedly performed, it can be very effective to evoke this sequence of operations using a short-hand: A function. In doing so, your code will gain enormously in power, convenience and elegance.

A function is created with `function(...) {...}`. As every R object, functions can be assigned to a symbol for future reference. (But functions can be used anonymously too.) The function's arguments (variables _and/or_ parameters) are given between round brackets in the function call, the operations to execute follow in curly brackets.

The example from above could look like that.

```{r}
fmol_to_abs <- function(x_in_fmol) {
  
  N_Avogrado = 6.022e23
  
  N_abs = x_in_fmol * N_Avogrado * 1e-15
  
  return(N_abs)
  
}

fmol_to_abs(10)     # calculate for x_in_fmol = 10
fmol_to_abs(N_fmol) # calculate for x_in_fmol = N_fmol, which is still assigned the value 2
```

Neither `N_abs` nor `N_Avogrado` have been added as objects to the ‘Global Environment’. They are defined but locally _within_ the function.

```{r}
ls()
```

Note that even if we had assigned `N_Avogrado = 1` in the ‘Global Environment’, our function _would prioritize_ the ‘locally’ defined symbol `N_Avogrado`, which is `6.022e23`. Only if we had _not_ defined this variable _inside_ the function, R would start looking for it outside the function. This is called **‘lexical scoping’**. 

If the last statement in a function is an expression (instead of an assignment), it is implicitly returned. So, we can simplify.

```{r}
fmol_to_abs <- function(x_in_fmol) {
  
  N_Avogrado = 6.022e23
  
  x_in_fmol * N_Avogrado * 1e-15
  
}

fmol_to_abs(N_fmol)
```

It is possible to make the return value of the function dependend on additional parameters, so that the function can be applied to different, yet similar problems. If a function takes a number of parameters, it can be convenient to assign default values to some arguments.

```{r}
mol_to_abs <- function(x, prefix = "f") {
  
  prefix_meaning = c("a" = 1e-18, # “c” is a function that combines values into a “list”
                     "f" = 1e-15, 
                     "p" = 1e-12,
                     "n" = 1e-09,
                     "µ" = 1e-06,
                     "m" = 1e-03)
  
  x * 6.022e23 * prefix_meaning[[prefix]]
  
}
```

If an argument with a default is omitted from the function call, R will silently use the default value.

```{r eval=FALSE}
mol_to_abs(20)               # amount in fmol to absolute numbers
mol_to_abs(20, prefix = "µ") # amount in µmol to absolute numbers
```

Unnamed arguments are interpreted in the positional order given in the function's definition. Named arguments can be provided in any order. The following expressions thus return the same value.

```{r eval=FALSE}
mol_to_abs(20, "µ")
mol_to_abs(prefix = "µ", x = 20)
```

## Packages

Not all functions are loaded by default into R's memory. Also, not all functions required for a task might even be included in the “standard distribution” of R. Therefore, users can write, deposit and load extensions to the R software, which come as bundles of functions and/or other objects in so-called ‘packages’. 

### Installing a New Package

R packages can be found anywhere on the internet, but most reliably, they are obtained from the ‘Comprehensive R Archive Network’ ([CRAN](https://cran.rstudio.com)), the main repository of contriubted R packages. 

This is how to install a CRAN package that is presumably not on your machine from CRAN, e.g. the `magrittr` package.

```{r eval=FALSE}
install.packages("magrittr")
```

To uninstall a package, use `remove.packages(package_name)`. (You will rarely need this.)

### Loading a Package

Simply installing a package is not sufficient to call its functions in the current R session already. Only a few packages are loaded by default when R starts up. This keeps the memory free from unnecessary assignments.

To load an additional package into R's memory, use

```{r eval=FALSE}
library("magrittr")
```

All of the package's function will now be known in the current R session. More specifically, their **‘namespace’** has been ‘attached’ to where R looks for object names, the so-called ‘search path’. To check which namespaces are currently known, type `loadedNamespaces()`.

In RStudio, the object in the current ‘search path’ can be browsed by toggling the ‘Global Environment’ drop-down menu.

To remove a namespace from the current session, type `unloadNamespace(namespace_name)` with the respective package name. (You will rarely do this.)

Note that some packages are automatically installed and silently loaded/attached when another package needs some of its functions. You cannot unload such namespaces unless you unload all the dependend packages first. 

### Referring to Functions from a Specific Package

If you intend to use only a single function from a package (and do not want to attach the entire namespace), or if there are functions with the same name from different packages, you must be more explicit and prefix the function call with the package name separated by `::`, e.g. `magrittr::divide_by(x, y)`. 

> When you load a package with conflicting names, R will print a warning message.

## Getting Help

Needless to say, there are probably very few questions and problems that another person has not yet dared asking on the internet. Some suggestions are listed on [https://www.r-project.org/help.html](https://www.r-project.org/help.html#r-help-on-the-internet).

However, if you want to work offline and/or get to know the how and why, you can do it with R!

### Checking the Source Code

To ‘see’ the code that a function executes, just type its name without parentheses into the R console.

```{r}
mol_to_abs
```

Most R functions (except for some primitve ones) are written in R and therefore no different from user-defined ones.

### Checking the Documentation

Sometimes, just looking at a function's code can be rather enigmatic. Therefore, all (all!) functions in R come with an extensive **documentation**. Here is how you get there.

1. _If you know the command's name,_ the following approach will open the appropriate documentation. As an example, let's examine the help for the `sum(...)` function.

    ```{r eval=FALSE}
help(sum)
help("sum")
?sum
    ```

    Typically, the help page includes the following sections:

    * The ‘Description’ states briefly what the function does.
    * The ‘Usage’ section specifies all arguments (variables and parameters) in positional order with their default values; here `...` represents any arbitrary number of comma-separated (unquoted) arguments.
    * ‘Arguments’ enlists each of the arguments and the data type expected.
    * ‘Details’ is a (often long) section that describes the function's behaviour, its intended usage, and other important details on how the function operates.
    * ‘Value’ explicitly describes the return value of the function.
    * ‘Examples’ gives illustrative code that serves the better understanding the function's behavior.
    
    Here is some useful advice on how to read a help file from Patrick Burns' [Impatient R](https://www.burns-stat.com/documents/tutorials/impatient-r/#readhelp):
    
    1. Help files are not novels. Don't feel compelled to read them from start to finish!
    2. Focusing on the examples may be a good strategy---provided that they are good examples ...
    3. Try it out! It may not be wise to expect yourself to understand everything before you _use_ the function.

2. _If you do not know the exact name of the function you are looking for,_ but you know what the function _should_ be able to do for you, one of the following commands will be helpful.

    ```{r eval=FALSE}
help.search("histogram")
??histogram
    ```

Many packages include **‘vignettes’**, which are documents to illustrate the facilities of the particular package. You can discover vignettes by accessing the index of the help page for a package or e.g. via `browseVignettes(package = "magrittr")`.

Some packages also include extended **code demonstrations**, called ‘demos’. Access e.g. via `demo(package = "stats")`.

## Errors and Warnings

R produces errors and warnings. Both write a message to your console. The difference is that errors halt the execution of the command, but warnings don't.

Although there are endless ways of getting an error, R is usually very good in pointing you to its source. 

Common sources of **errors** include

* missing commas,
* typing unmatched parentheses or trying to match parentheses of different kinds,
* misspelling object names/symbols or their capitalization, or
* failing to attach the relevant packages to the current namespace.

**Warnings** are not as serious as an error. 

However, ignoring a warning can be very serious if it is suggesting that the answer you got was bogus.

> Always (always!) _read_ the warning messages to check if they indicate a real problem or not. 

## Summary

In this introduction you have learned

* how to ask R to execute a command using the command line prompt,
* how to create an object, consisting of a value and a name,
* how to check which objects are currently in R's memory, and
* how to remove them,
* what a function does,
* what a package is,
* how to find help on functions and packages, and
* the difference between errors and warnings.

## Quick Questions

1. What do `#`, `%>%`, and `->` mean?
2. How do you type `{`, `}`, `[`, and `]` on your keyboard?
3. Check out the documentation for `rm`. Can you figure out what `rm(list = ls())` will do? Do it!