---
title: "Working with Strings"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This introduction will be on objects of type `character`. This is all text which appears between quotation marks (`""` or `''`) and is therefore not executed by R. In informatics, such text objects are called ‘strings’.

Note how to get the “length” of a string in R, i.e. the number of characters.

```{r}
# number of characters
nchar("apple")
# number of elements in the character vector
length("apple")

fruits <- c("apple", "banana", "lemon")
# number of elements in the character vector
length(fruits)
# number of characters in each element of the vector vs length of each element in the vector
sapply(fruits, nchar); sapply(fruits, length)
```

## Combining Strings

The `paste(...)` function can be used to contract strings (e.g. two `character` objects) into a single string (one `charachter` object).

```{r}
paste("banana", "cake")
```

If you supply two vectors, the shorter is repeated over the length of the longer one. Vectorization will make a vector from the vectors.

```{r}
paste(c("banana", "chocolate", "doughnut"), "cake")
```

Here is an illustration of the agruments `sep=...` and `collapse=...`.

```{r}
x <- paste(c("A", "B", "C", "D"), c(1, 2), sep = "_")
x
length(x) # a vector with four elements

x <- paste(c("A", "B", "C", "D"), c(1, 2), sep = "_", collapse = " and ")
x
length(x) # a vector with one element
```

The function `paste(..., sep = "")` which glues all arguments without spaces, has a shortcut, `paste0(...)`.

## Splitting Strings

Strings can be split after a specific character with base R's `strsplit()` function.

```{r}
strsplit(x = "one,2;three,4", split = ",")
```

An equivalent, but slightly more powerful function is implemented in the `stringr` package. It is vectorized over both arguments, `string` and `pattern`.

```{r}
stringr::str_split(string = "one,2;three,4", pattern = c(",", ";"))
```


## Subsetting Strings

To extract a substring from a string, base R has a function called `substr()`. For the sake of consistency however, we will focus the equivalent functions of the `stringr` package, which is part of the `tidyverse`.

Subsetting is done by position. Negative indices (in `stringr`, not available in base R) count from the end.

```{r}
library(stringr)

fruits <- c("apple", "banana", "lemon")

# the second letter
str_sub(fruits, start = 2, end = 2)
# the third to second to last letters
str_sub(fruits, start = 3, end = -2)
```

You can use `str_sub()` to modify strings too.

```{r}
str_sub(fruits, start = 3, end = -2) <- "x"
fruits
```

To remove whitespace from the ends of a string, there is `str_trim()`. 

```{r}
str_trim("   chitchat  ")
```

## Finding and Replacing Strings in Strings

In its easiest implementation, finding and replacing strings works as follows with the `stringr` package.

```{r}
str_replace(string  = "One small step for a man, one giant leap for mankind!",
            pattern = "man",
            replacement = "mouse")
```

Note that `str_replace()` replaces only the _first_ occurence of a the ‘pattern’. If you want to replace _all_ occurences, use `str_replace_all()`.

```{r}
str_replace_all(string  = "One small step for a man, one giant leap for mankind!",
                pattern = "man",
                replacement = "mouse")
```

## Regular Expressions

In many cases we would be glad to find and replace not only words exactly matching our query, but also words that are constructed in a specific manner (that have a specific ‘pattern’). The ‘language’ in which such patterns are given is called ‘[regular expression](https://www.regular-expressions.info/tutorial.html)’ and is by far not limited to R, but applicable to many programming languages. 

This is a brief overview with some examples modified from the [`stringr` vignette on regular expressions](https://stringr.tidyverse.org/articles/regular-expressions.html). 

In almost all `stringr` functions, you can use regular expressions as `pattern=...`.

```{r}
str_replace_all(string  = "One small step for a man, one giant leap for mankind!",
                pattern = "step|leap", # match either ‘step’ or ‘leap’
                replacement = "piece of cake")
```

In the next examples, we will use the `str_extract_all()` function, which will return the matching substrings of each element in the `character` vector provided as `string=...`. Since we have only one element in this vector, we use index `[[1]]` to print out the matches for the first (and only) element.

```{r}
str_extract_all(string  = "One small step for a man, one giant leap for mankind!", # one element only
                pattern = "step|leap")[[1]]
str_extract_all(string  = c("One small step", "one giant leap"), # character vector with two elements
                pattern = "step|leap")
```

Regular expressions can have placeholder symbols. First, `.` will match any character (except a newline). 

```{r}
str_extract_all(string  = "I keep pressing the space bar, but I'm still on planet Earth.",
                pattern = ".ar")[[1]]
str_extract_all(string  = "I keep pressing the space bar, but I'm still on planet Earth.",
                pattern = ".ar.")[[1]]
```

If you need to match a point, you need to use an ‘escape’ to tell the regular expression not to use the special meaning of `.`, but its literal. 

> The ‘escape’ sign for regular expressions is `\`, so you would need `\.`. However, `\` is also the escape sign for R. So, in order to _perserve_ the escape sign for the regular expression, you need to escape the escape sign as well. (Sounds confusing, I know.) 

You will need `\\.` to match a fullstop literally.

```{r}
str_extract_all(string  = "If you obey all the rules, you miss all the fun.",
                pattern = "fun\\.")[[1]]
```

### Quantifiers

To control how many times a pattern matches, you can use the repetition operators.

| operator | repetitions |
|----------|-------------|
| `?` | 0 or 1 |
| `+` | 1 or more |
| `*` | 0 or more |

```{r}
x <- "1899 or MDCCCXCIX is the last year of the 1890s decade."
str_extract(x, "CC?")
str_extract(x, "CC+")
str_extract_all(x, "CX*")[[1]]
```

You can also specify the number of matches using curly brackets.

| operator | repetitions |
|----------|-------------|
| `{n}` | exactly _n_ times |
| `{n,}` | _n_ or more times |
| `{n,m}` | between _n_ and _m_ |

```{r}
str_extract_all(x, "C{1,2}")[[1]]
```

By default, these matches are ‘greedy’. They will match the longest string possible. To make them match the shortest string possible, put a `?` after them.

| operator | repetitions |
|----------|-------------|
| `??` | 0 or 1, prefer 0 |
| `+?` | 1 or more, match as few as possible |
| `*?` | 0 or more, match as few as possible |
| `{n,}?` | _n_ or more, match as few as possible |
| `{n,m}?` | match at least _n_, but then as few as possible |

```{r}
str_extract(x, c("C{2,3}", "C{2,3}?"))
```

### Special Characters

Escapes allow to specify individual chracters that are otherwise hard to type.

* `\s` matches any whitespace. This includes horizontal tabulation signs (`\t`), line feeds (`\n`), carriage returns (`\r`), as well as a variatey of space characters and other separators defined in Unicode. 

    The complement `\S` matches any non-whitespace character.

    ```{r}
str_replace_all("If   there is\tno  space in 
                
                space,\nI will leave.  ", "\\s+", " ")
    ```

Escapes can also be useful to specify entire classes of characters.

* `\d` matches any digit (0, 1, 2, ..., 9). The complement, `\D`, matches any character that is not a decimal digit.

    ```{r}
str_extract_all("There are 214 moons in the solar system.", "\\d+")[[1]]
str_extract_all(x, "\\d+")[[1]]
    ```

* `\w` matches any ‘word’ character. Words consist of alphabetic characters (including diacritics etc.) and decimal numbers. 

    The complement `\W` matches any non-word character.

    ```{r}
str_extract_all("Don't shoot for the moon!", "\\w+")[[1]]
str_extract_all("Don't shoot for the moon!", "\\W+")[[1]]
    ```

* `\b` matches the transition between word and non-word characters. (Rarely needed.)

You can specify your own _character classes_ using `[]`:

* `[abc]` matches a, b, or c,
* `[a-z]` matches every character between a and z (case-sensitive),
* `[^abc]` matches anything except a, b, or c.

There are a number of pre-built character classes:

* `[:punct:]` punctuation,
* `[:alpha:]` letters,
* `[:lower:]` lowercase letters,
* `[:upper:]` uppercase letters,
* `[:digit:]` digits,
* `[:alnum:]` letters and numbers,
* `[:cntrl:]` control chacracters,
* `[:graph:]` letters, numbers, and punctuation,
* `[:print:]` letters, numbers, punctuation, and whitespace,
* `[:space:]` space characters (same as `\s`),
* `[:blank:]` space and tab.

### Anchors

In some cases, you want to match a pattern only if it occurs at the start or the end of a line. There are

* `^` to match the start of a line, and
* `$` to match the end of the line.

```{r}
str_extract_all("papaya or banana", "[yn]a")[[1]]
str_extract_all("papaya or banana", "[yn]a$")[[1]]
str_extract_all("papaya or banana", "[yn]a\\b")[[1]]
```

To remember which one is which, this [mnemonic](https://twitter.com/emisshula/status/323863393167613953) might help you: ‘If you begin with power (`^`), you end up with money (`$`).’

## Learning Objectives

In this introduction you should have learned

* how to combine and split strings,
* how to subset strings based on character indices,
* how to replace words in strings, and
* that you can use regular expressions to formulate advanced string queries.

Quick questions.

1. What is the result of `strsplit(paste0("AG", "CGT", "A", "TGCT"), "GC")`?
2. Suppose you have the vector `c("ÁlM", "AnJ", "DaS", "DmS", "JaW", "ShP", "SuB")`. Using regular expressions, how do you identify all words that contain
    1. the lowercase ‘a’,
    2. ‘a’ or ‘A’, 
    3. any non-Latin character, 
    4. more than two consonants in a row,
    5. end (or start) with ‘S’? 
