---
title: "Selecting and Manipulating Data"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

plate_data <- readRDS("part_11-plate_data.rds")
```

In this section, you will learn how to select specific entries (rows) based on their content from a table. Also, you will learn how to retrieve specific variables (columns) from the table and/or add new variables based on exisiting values to the data.

Having combined and tidied the plate data, we now have an object `plate_data` containing

```{r echo=FALSE}
plate_data %>% 
  head(7) %>% 
  knitr::kable(format = "markdown", align = "c")
```

_(table abridged)_

## Manipulating Rows (“Cases”)

We have already seen how to add cases using `add_row(...)`. Let's see how to arrange and extract rows based on their content.

### Arranging Rows

Tables can be sorted by specific columns. This is rather a convenience for the user than a necessity to perform further manipulations.

```{r eval=FALSE}
plate_data %>% arrange(concentration, intensity)
plate_data %>% arrange(desc(concentration), intensity)
```

### Extracting Rows

These functions return a subset of rows as a new table.

```{r eval=FALSE}
# extract rows that meet criteria
plate_data %>% filter(sample_id == "treatment_B")
plate_data %>% filter(replicate_id == "rep.B" & concentration > 50)
plate_data %>% filter(intensity * concentration > 1e4)

plate_data %>% filter(intensity >= 1 & intensity <= 3)
plate_data %>% filter(between(intensity, 1, 3)) # same, shorter

plate_data %>% filter(near(intensity, 10, 0.5)) # safe for floating point numbers

# remove duplicate rows
plate_data %>% distinct(sample_id, concentration)
```

One can also select (randomly) a number of cases.

```{r eval=FALSE}
# randomly extract N rows without replacement
plate_data %>% sample_n(5, replace = FALSE)
# randomly extract % rows with replacement
plate_data %>% sample_frac(.1, replace = TRUE)

# select rows by position
plate_data %>% slice(4:6)

# order and select top N entries
plate_data %>% top_n(5, intensity)
```

## Manipulating Columns (“Variables”)

It might not be obvious at this point, but most power of the `tidyverse` is taken from its ability to operate on existing columns and create new columns.

### Selecting Columns

Many functions will require a proper selection of columns, e.g. during data tidying. This is why we focus here on how to specify a column selection before we move on to more illustrative examples.

> The `tidyverse` makes extensive use of ‘quasiquotation’. This means that you can refer e.g. to column names (which are usually of type `character`) without using quotation marks (`""` or `''`). But beware, some arguments must be `character` vectors. These must be quoted.

* To select a column, you can always **type its name between quotation marks**. If you want to specify multiple columns, use a `character` vector with the column names.

* Most functions will accept **unquoted column names**. You don't need a vector or anything like that then.

    Typically, this is indicated with `...` in the ‘Usage’ section of the Documentation. 

* In addition, there are **helper functions** to select multiple columns based on their position or based on their name.
    
    They are documented in `?select_helpers`.
    
    | operator/function | selection |
    |-------------------|--------|
    | `everything` | all columns |
    | `last_col` | last column |
    | `one_of` | these column names as character vector |
    | `starts_with` | column names starting with this prefix literally |
    | `ends_with` | column names ending with this suffix literally |
    | `contains` | column names containing this string literally |
    | `matches` | column names matching the regular expression |
    | `:` | all columns between |
    | `-` | all columns except |
    
    For ‘scoped’ summarizing (`summarize_at`) and mutating verbs (`mutate_at`), you will need to wrap the helper statement with `vars(...)`.
    
    For ‘scoped’ filtering verbs (`filter_all` or `filter_if`), there are additional variants called `all_vars(...)` and `any_vars(...)`.

### Extracting and Arranging Columns

Most frequently, you will need 

* `select` to (optionally rename and) extract columns as table, or
* `pull` to extract column values of _one_ column as a vector; which defaults to the last column is none is specified.

```{r eval=FALSE}
# return (last) column as vector
plate_data %>% pull
# return selected columns as table
plate_data %>% select(intensity)   # with quasiquotation
plate_data %>% select("intensity") # with explicit names
plate_data %>% select(starts_with("int"))
plate_data %>% select(ends_with("id"))
plate_data %>% select(c("sample_id", "replicate_id"))
```

There are the ‘scoped’ variants, 

* `select_all` to rename all columns (e.g. make uppercase),
* `select_at` to (optionally rename and) extract columns select with `vars(...)`, and
* `select_if` to (optionally rename and) extract columns that meet a certain property, (e.g. contain numeric values).

```{r eval=FALSE}
# select based e.g. on data type
plate_data %>% select_if(is.numeric)
# select columns and rename column names
plate_data %>% select_all(toupper)
plate_data %>% select_at(vars(ends_with("id")), ~str_c(., "iot"))
```

To rename columns only, there are `rename` variants of the above.

```{r eval=FALSE}
plate_data %>% rename("conc" = "concentration") # with explicit names
plate_data %>% rename(conc = concentration)     # with quasiquotation
plate_data %>% rename_at(vars(ends_with("id")), ~str_c(., c("ea", "iot")))
```

Less importantly, `select` can be used for re-arranging columns.

```{r eval=FALSE}
plate_data %>% select(concentration, ends_with("id"), intensity)
```

### Making New Variables

`mutate` and its ‘scoped’ variants will apply _vectorized_ functions on the selected columns and append a new column with the result of the transformation. 

`transmute` is a short-hand for `mutate` followed by `select` on the result.

```{r eval=FALSE}
# create a new column ...
plate_data %>% mutate(log10(concentration))
# ... with another name ...
plate_data %>% mutate(log_conc = log10(concentration))
# ... or in place
plate_data %>% mutate(concentration = log10(concentration))
# with subsequent selection
plate_data %>% transmute(log_conc = log10(concentration))
# multiple columns can be specified at the same time
plate_data %>% transmute(log_conc = log10(concentration), 
                         sqr_conc = concentration ** 2)
```

There are many useful functions in `dplyr` to add cumulative aggregates and ranks to the data. Have a look on this [cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf).

```{r eval=FALSE}
# cumulative sum 
plate_data %>% filter(!is.na(intensity)) %>% mutate(cumsum(concentration))
# break the input vector into N buckets based on rank
plate_data %>% filter(!is.na(intensity)) %>% mutate(bucket_no = ntile(intensity, 10))
# break the input vector continuously on rank
plate_data %>% filter(!is.na(intensity)) %>% mutate(percent_rank(intensity))
```

An interesting application is to replace values or create variables based on a combination of logical evaluations.

```{r eval=FALSE}
# replace an annoying value with NA
plate_data %>% mutate(concentration = na_if(concentration, 0))
# replace all missing values with a value
plate_data %>% mutate(replace_na(intensity, 0))

# general replacements
plate_data %>% 
  mutate(sample_id = recode(sample_id, control = "lemon", 
                            replicate_A = "banana",
                            replicate_B = "apple"))

# replace based on logical conditions
plate_data %>% filter(!is.na(intensity)) %>% 
  mutate(intensity = if_else(concentration == 100, intensity * 1e4, intensity))

# replace based on multiple conditions ...
plate_data %>% filter(!is.na(intensity)) %>% 
  mutate(intensity = case_when(
    concentration == 0   ~ intensity + 2e2,
    concentration == 1e2 ~ intensity * 1e4,
    # all other cases
    TRUE ~ intensity
  ))

# ... especially useful when you want to create a new variable based on a
# complex combination of existing variables
dplyr::starwars %>%
  select(name:mass, gender, species) %>%
  mutate(type = case_when(
    height > 200 | mass > 200 ~ "large",
    species == "Droid"        ~ "robot",
    TRUE                      ~ "other"
  ))
```

### New Variables from Non-Vectorized Functions



## Hands-On Exercise

1. Change `"rep.A"` and `"rep.B"` into `"replicate_3"` and `"replicate_4"` respectively, arrange by `sample_id` and `replicate_id` (descending), then store the result as `plate_data`.

```{r include=FALSE}
plate_data %>% 
  mutate(replicate_id = recode(replicate_id,
                               rep.A = "replicate_3",
                               rep.B = "replicate_4")) %>% 
  arrange(sample_id, desc(replicate_id)) -> plate_data
```

2. Subtract `control` from all treatments (without modifying the data set).

```{r include=FALSE}
plate_data %>% 
  # make long table wide
  pivot_wider(names_from = "sample_id", values_from = "intensity") %>% 
  # subtract control in place
  mutate_at(vars(starts_with("treatment")), funs(. - control))
```

```{r include=FALSE}
saveRDS(plate_data, file = "part_12-plate_data.rds")
 ```