---
title: "Getting Started---A Cheat Sheet"
author: "Benjamin Buchmuller"
output: 
  word_document:
    df_print: kable
    reference_docx: word_styles_references.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic Interactions

### Code and Comments

* `#` is the comment character
* `a = b` and `a <- b` or `b -> a` **assigns** the value of `b` to a variable named `a`
* `function_a %>% function_b` **pipes** the result of `function_a` to `function_b` as input

Object names cannot start with a number, some names are reserved.

### Functions and Packages

The basic idiom to define a function is `new_function <- function(arg1, arg2, ...) {...}`.

Packages

* include new functions for a specific task,
* are installed with `install.packages("some_package")`,
* are loaded with `library(some_package)`.

Functions from a package can be specifically pointed with `some_package::this_function(...)`.

### Getting Help

If you know the command's name use `?command`.

If you don't know the name, but a keyword, use `??keyword`.

## Data Types

| data type | stores | example |
|-----------|--------|---------|
| `double` | floating point numbers | `3.141, 1.8e-3` |
| `integer` | integers | `-5L, 1L, 2L, 3L` |
| `character` | strings | `"banana", "apple", "melon"` |
| `logical` | boolean values | `TRUE, FALSE` |
| `list` | sets with different data types | `list(3, "apple")` |

```{r eval=FALSE}
typeof(obj)     # data type of obj
class(obj)      # class of obj; will tell functions how treat obj
attributes(obj) # metadata of obj; a named list
```

## Vectors

> Vectors contain elements of one single data type only!

```{r eval=FALSE}
length(obj)     # number of elements in obj
c(obj_a, obj_b) # concatenate obj_a and obj_b; may involve coercion 
names(obj)      # names of each element in obj; if set, a character vector 
```

### Numerical Vectors

```{r eval=FALSE}
1:10            # vector of type integer
seq(1, 10, 0.5) # vector of type double
```

### Logical Vectors

```{r eval=FALSE}
1:4 > 2         # vector of type logical
# negation of logical vectors
!(1:4 > 2)
# boolean operations 
c(TRUE, FALSE) & c(TRUE, TRUE) # AND : c(TRUE, FALSE)
c(TRUE, FALSE) | c(TRUE, TRUE) # OR  : c(TRUE, TRUE)
# boolean queries
any(...) # at least one element TRUE?
all(...) # all elements TRUE?
```

### Indexing Vectors

```{r eval=FALSE}
# index by position
noble_gases[c(1, 3, 5)]
# index by exclusion
noble_gases[c(-1, -3)]
# index by name
noble_gases[c("Argon", "Radon")]
# index by logical
noble_gases[noble_gases < 20]
```

### Applying Functions to Vectors

Functions that take _one_ vector as input are applied `fun(obj)`. 

Functions that take _two_ vectors as input are applied `fun(obj_a, obj_b)`, this includes arithmetic operations `obj_a + obj_b`.

* If `obj_a` and `obj_b` have the same number of elements, `fun` is applied by pairs,
* if `length(obj_a)` is a multiple of `length(obj_b)`, `obj_b` is recycled along `obj_a` to make pairs,
* else, there will be a result with a warning.

To apply a function to vectors inside a `list`, the basic idiom for is `lapply(obj, fun)`.

One can apply even complicated functions using `lapply(obj, function(i) ...)`.

> Using vectorization is preferred over `for`-loops!

## Working with Strings

```{r eval=FALSE}
# number of characters
nchar(chr)

# glue together chr_a and chr_b by pairs of elements
paste(chr_a, chr_b)  # with sep=...
paste0(chr_a, chr_b) # with no separator
# glue together chr_a and chr_b by pairs of elements, then combine in a single string
paste(chr_a, chr_b, collapse = ...)  # with sep=... between elements
paste0(chr_a, chr_b, collapse = ...) # with no separator
# split chr_a at each occurence of chr_b
stringr::str_split(chr_a, pattern = chr_b)

# extract the substring from the fourth to the eighths character
stringr::str_sub(chr, start = 4, end = 8)

# find first chr_b (can be regex) in chr_a
stringr::str_extract(chr_a, pattern = chr_b)
# find all chr_b (can be regex) in chr_a
stringr::str_extract_all(chr_a, pattern = chr_b)

# replace first chr_b (can be regex) in chr_a with chr_c
stringr::str_replace(chr_a, pattern = chr_b, replacement = chr_c)
# replace all chr_b (can be regex) in chr_a with chr_c
stringr::str_replace_all(chr_a, pattern = chr_b, replacement = chr_c)
```

### Regular Expressions

#### Character classes.

| operator | meaning |
|----------|---------|
| `.` | any character (except `\n`) |
| `\s` | any whitespace |
| `\t` | horizontal tabulation signs |
| `\n` or `\r` | line feeds or carriage returns |
| `\d` | any digit (0, 1, 2, ..., 9) |
| `\w` | any alphabetic and decimal numbers | 
| `[abc]` | a, b, or c |
| `[a-c]` | every character between a and z (case-sensitive) |
| `[^abc]` | anything except a, b, or c | 

#### Quantifiers.

| operator | repetitions |
|----------|-------------|
| `?` | 0 or 1 |
| `+` | 1 or more |
| `*` | 0 or more |
| `{n}` | exactly _n_ times |
| `{n,}` | _n_ or more times |
| `{n,m}` | between _n_ and _m_ |
| `...?` | (make not greedy) |

#### Anchors.

| operator | position |
|----------|----------|
| `^` | start of a line |
| `$` | end of a line |