---
title: "Grouping and Summarizing Data"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

plate_data <- readRDS("part_11-plate_data.rds")
```

The important object you will learn in this section is how to summarize (large) data sets. Typically, the data is first split into several groups defined by a categorical variable (‘factor’) and then some sort of summary statistics are applied to each group.

## Making Groups

`group_by(...)` specifies the columns containing the categorical variables on which you want to split your data set. Note that it is _not_ necessary (but sometimes useful) to convert the column into a `factor`.

Grouping does not change how the data looks like, apart from listing how it's grouped.

```{r}
plate_data %>% group_by(sample_id)
```

To check how many entries there are per group, there is `group_data()`.

```{r}
plate_data %>% group_by(sample_id, replicate_id) %>% group_data()
```

We could also use some ‘scoped’ variants such as `group_by_at(...)`, `group_by_if(...)` or `group_by_all(...)` to select the columns on which to group.

```{r eval=FALSE}
plate_data %>% group_by_at(vars(ends_with("id"))) # group by "sample_id" and "replicate_id"
```

By default, `group_by(...)` overrides existing grouping of the data set. Use `group_by(..., add = TRUE)` to append.

To remove the grouping, use `ungroup()`.

## Summarizing Groups

A common summary of continuous data is to give their average value and the spread. Given `plate_data`, let's calculate the average intensity for each `sample_id` and `concentration`.

```{r}
plate_data %>% 
  group_by(sample_id, concentration) %>% 
  # calculate mean and sd
  summarize(mean = mean(intensity, na.rm = TRUE), 
            sd = sd(intensity, na.rm = TRUE), 
            # count the rows that are not NA
            N  = sum(!is.na(intensity)))
```

Suspiciously, the standard deviation of the samples with four replicates (`replicate_1`, `replicate_2`, `replicate_3` and `replicate_4`) are higher than the others. We shall keep this in mind as there might be something ‘wrong’ with one of the replicates.

For categorical data, we are more likely to be interested in the number of observations, `n()`, or the number of unique values a variable takes within a group, `n_distinct()`. 

There are many more summary functions. Have a look on this [cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf).

## Comparing Groups

Statistics is a powerful way to quantify how different supposedly different groups are from each other. To make uni- and multivariate comparisons or to fit models through data based on groups needs just some more tweaks in the `tidyverse`.

### Student's _t_ Test

Here, we will use a built-in data set in R, `reshape2::tips`, which contains a record on the tips a waiter received over several months in restaurant.

```{r}
reshape2::tips %>% as_tibble()
```

For example, we might be curious whether male or female guests give higher tips. In base R, we could go with this one-liner.

```{r}
t.test(tip ~ sex, data = reshape2::tips)
```

However, if we want to perform this test by weekday, we are left with subsetting the data (manually) by day and apply the following on each subset.

```{r}
reshape2::tips[which(reshape2::tips$day == "Fri"), ] -> tips.Fri
t.test(tip ~ sex, data = tips.Fri)
```
```{r include=FALSE}
rm(tips.Fri)
```

We would certainly prefer a neat table with the parameters of the fit organized as columns and the rows specifying the grouping variables such as the weekday. In this sense, a vectorized approach.

Let's look how to perform a Student's _t_ test within the `tidyverse`. 

We will use `group_modify(...)`, a summary function for cases when `summarize(...)` is too limited. It takes a function as argument, which operates on and returns back a `data.frame`. In this case, the function will be given the (four) different `data.frame` objects, one for each group (day). 

> Since R's `t.test` does not return a `data.frame` by default, we wrap it with `~ broom::tidy(...)`. This function makes a `data.frame` containing the parameters (e.g. coefficients and slopes) of the model. 

> Note: The `~` in front of `broom::tidy(...)` is mandatory for reasons beyond the scope of this tutorial at this point.

Within `group_modify`, we can refer to the current subset with `.` (or `.x`) and the value of the grouping variable with `.y`.

```{r}
reshape2::tips %>% 
  group_by(day) %>%
  group_modify(~ broom::tidy(t.test(tip ~ sex, data = .)))
```

A little down-side of this nicely general approach is that column names may become rather enigmatic. This is because they need to apply to very different cases in which `broom::tidy(...)` is used. 

Here, the columns named `estimate`, `estimate1` and `estimate2` refer to the mean difference between both groups, the mean of the first group (here: ‘Female’) and the mean of the second group (here: ‘Male’) respectively.

### Student's _t_ Test with Multiple Comparisons

If more than two groups are compared, we should adjust the p-values for the additional comparisons. The function is a little bit misleadingly named `pairwise.t.test` and can be used for paired and unpaired comparisons between multiple groups.

First, the same as above.

```{r}
reshape2::tips %>% 
  group_by(day) %>% 
  group_modify(~ broom::tidy(pairwise.t.test(
    x = .$tip, # the column with the values to compare
    g = .$sex, # the column from which the different groups are taken
    p.adjust.method = "none", pool.sd = FALSE)))
```

Let's group by sex and compare the tips for each weekday without applying a correction.

```{r}
reshape2::tips %>% 
  group_by(sex) %>% 
  group_modify(~ broom::tidy(pairwise.t.test(
    x = .$tip, g = .$day, p.adjust.method = "none", pool.sd = FALSE)))
```

However, when properly done, the ‘significant’ difference for female payers between Thursdays and Sundays vanishes.

```{r eval=FALSE}
reshape2::tips %>% 
  group_by(sex) %>% 
  group_modify(~ broom::tidy(pairwise.t.test(
    x = .$tip, g = .$day, p.adjust.method = "hochberg", pool.sd = FALSE)))
```

This approach can also be applied to other hypothesis testing functions such as `cor.test`, `wilcox.test`, `chisq.test` etc.

### Linear Regression and Beyond

Another example on the same lines is a linear regression by group. Let's try to describe the `intensity` observed in `plate_data` as a function of the `concentration`. Such a model is fit in R with the linear model `lm(...)` call.

```{r}
plate_data %>% 
  group_by(sample_id) %>% 
  group_modify(~ broom::tidy(lm(intensity ~ log10(concentration + 1), data = .)))
```

So far, we have used `broom::tidy` to get the parameters of the model. If we are interested in the statistical summary of the overall model, we use `broom::glance`.

```{r}
plate_data %>% 
  group_by(sample_id) %>% 
  group_modify(~ broom::glance(lm(intensity ~ log10(concentration + 1), data = .)))
```

All `broom` functions apply equally well to the output of generalized linear and non-linear models. 