---
title: "Importing and Tidying Data"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

If you have not done so yet, please attach the `tidyverse`. 

```{r message=FALSE}
library(tidyverse)
```

## Importing Data from Delimited Files

A common standard to store tabulated data is as comma-separated values. Typically such files end with the extension `.csv`. However, the delimiters to represent the start of a new column could be any character such as a semicolon or a white space (e.g. tab, blank).

### Importing Data from a Single File

The data of the plate mentioned in [this session's introduction](part_10-working_with_tables.html) was saved as a comma-delimited file called ‘plate_1.csv’. You can find it in the sub-folder ‘part_10-working_with_tables_files’. We are going to import this file using `readr::read_csv(file)`. This function will try to guess the data type stored in each column.

> In the file path, `.` refers to the current working directory, `getwd()`, so that we can use relative paths.

```{r}
plate_1 <- read_csv(file = "./part_10-working_with_tables_files/plate_1.csv")
plate_1
```

The column names have been taken from the first line of the file (since they were text). However, as we do know the actual names, we should use them. They can be set either using `colnames(plate_1)<-` or during import. In the latter case, we can (and should) skip the first line of the file since it's obsolete.

```{r}
plate_1 <- read_csv(file = "./part_10-working_with_tables_files/plate_1.csv",
                    col_names = str_c("conc_", c(1:4, 0)), skip = 1)
```

The data we are typically dealing with is much too complex to rely on a single row name only. To be useful, each parameter or variable should be stored in a separate column. 

So, let's add one column specifying the sample type and one column specifying the replicate number.

```{r}
plate_1 <- plate_1 %>% 
  add_column(sample_id = rep(c("control", "treatment_A"), each = 2),
             replicate_id = rep(str_c("replicate_", c(1, 2)), times = 2),
             # insert before first column; default after last column
             .before = 1)
plate_1
```

If you had (for some reason) rownames in a `data.frame` that you want to keep in the `tidyverse`, use `dplyr::rownames_to_column()`. 

### Combining Data from Multiple Files with `row_bind`

A nice feature of using a programmatic approach for data analysis is that it allows retrieving (systematically) many files from multiple locations on the harddrive (e.g. reflecting multiple experiments) and to analyze them either in identical ways one by one, or to combine them for analysis. 

Certainly, we could import each file individually …

```{r eval=FALSE}
plate_2 <- read_csv(file = "./part_10-working_with_tables_files/plate_2.csv")
plate_3 <- read_csv(file = "./part_10-working_with_tables_files/plate_3.csv")
```

However, if we have many files to analyse, we get easily lost. A more amenable approach is using `list.files(...)` to create a list (`character` vector) of all files in a directory and apply regular expressions to specify the files we would like to include.

```{r}
plate_files.paths <- list.files(path = "./part_10-working_with_tables_files",
                                pattern = "plate_[^1]", full.names = TRUE)

plate_files <- sapply(plate_files.paths, read_csv, simplify = FALSE, USE.NAMES = TRUE)
plate_files
```

The `plate_files` object is a `list` with two `tibble` objects (`dplyr`'s way of `data.frame`). Apparently, two columns have been swapped in `plate_3`, which could cause problems if we just pasted them mindlessly below each other. 

Luckily, R pays attention for us.

```{r}
plate_files %>% bind_rows(.id = "file")
```

We have kept track of the file names (or even file paths) by specifying an `.id` column called `"file"` from the names of the `tibble` `list`. If need be, we could extract metadata stored in the file name or file path by creating new columns. (Details in the next session.)

```{r}
plate_files %>% bind_rows(.id = "file") %>% mutate(file = str_extract(file, "plate_[0-9]+"))
```

For the moment, let's combine the new data with `plate_1` without the file paths and call the object `plate_data`.

```{r}
plate_data <- bind_rows(plate_1, plate_files)

# remove objects no longer needed
rm(plate_files, plate_files.paths)
```

### Combining Data with `join`

We have seen that the command to combine rows is `dpylr::bind_rows(...)` (and `rbind(...)` in base R). Likewise, there is `dplyr::bind_cols(...)` (or `cbind(...)` in base R) to combine columns. However, we must be absolutely sure that the rows align!

```{r eval=FALSE}
cbind(
  # half plate with conc_1 measurments
  plate_data %>% select(sample_id, replicate_id, conc_1),
  # half plate with conc_2 measurements
  plate_data %>% select(sample_id, replicate_id, conc_2)
)
```

In the following case, this fails. (Don't care too much about how to create the randomized halfs.)

```{r include=FALSE}
set.seed(14493)
```
```{r}
# half plate with conc_1 measurments
half_1 <- plate_data %>%
            select(sample_id, replicate_id, conc_1)
# half plate with conc_2 measurements
half_2 <- plate_data %>% sample_n(nrow(.)) %>% 
            select(sample_id, replicate_id, conc_2)

# this will be wrong
bind_cols(half_1, half_2)
```

To properly merge `half_1` and `half_2`, we need to merge the tables based on the colums `sample_id` and `replicate_id`. This is what `dplyr::inner_join(x, y)` does.

```{r}
# this will be correct
inner_join(half_1, half_2, by = c("sample_id", "replicate_id"))
```

If you want to keep rows that are in one table, but not in another, there are also the functions `dpylr::left_join(x, y)` and `dplyr::right_join(x, y)`. Having both options is useful when you think of piping.

```{r include=FALSE}
save(half_1, half_2, file = "part_11-half_plates.RData")
```
```{r}
# remove objects no longer needed
rm(half_1, half_2)
```

### Importing Data from Excel

To read data from Excel, you can use `readxl::read_excel(file)`. The `readxl` is a package of the `tidyverse`.

Here is an example with `plate_1.xlsx`, which contains the same data put somewhere on the worksheet. By default, `readxl::read_excel(file)` uses the smallest rectangle that contains the non-empty cells.

```{r}
readxl::read_excel("./part_10-working_with_tables_files/plate_1.xlsx")
```

The blank column headers will get a new, unique name (e.g. `...1`, `...2` etc.) and the content of merged cells will be assigned to the top-most left cell of the area.

However, I would disencourage to rely on routine importing data from Excel workbooks since `.xls` and `.xlsx` are neither open (i.e. disclosed), nor standardized file formats. This means they can be changed by Microsoft anytime and may consequently not be (immediately) supported by other software.

Another problem you might be facing is that within Excel, it is tempting to put multiple tables with different data side-by-side in a single worksheet. To import such worksheets  will require a lot of polishing from your side after.

> If you plan to import from Excel, keep it simple and put one set of data per sheet. Ideally, export as `.csv`.

### Creating a `data.frame` from Scratch

Sometimes it's necessary to quickly make a `data.frame` or related objects.

This base R approach works _always_.

```{r}
set.seed(181096)
```
```{r}
# simulate some data
status <- factor(rep(c("smoker", "non-smoker"), each = 50))
risk_non_smoker <- rnorm(50)
risk_smoker <- 2 + risk_non_smoker

# create a data.frame by column
smoking_risk <- data.frame(status = status, risk = c(risk_smoker, risk_non_smoker))

head(smoking_risk)

# create a quick plot
plot(risk ~ status, data = smoking_risk)
```

Remember that a `data.frame` is basically a `list` of vectors (of identical length) that can contain a different type of data each. This is why tables (if not imported) are created by column from vectors. 

The `tidyverse` provides and uses a special case of the `data.frame` class, which is the `tibble`. It has a couple of advantages (see `?tibble::tibble`).

`tibble` objects are created like `data.frame` objects.

```{r}
smoking_risk <- tibble(status = status, risk = c(risk_smoker, risk_non_smoker))

smoking_risk # no head() call required, prints only top rows
```

If for some reason you want to create a `tibble` by rows, use `tibble::tribble`.

```{r}
tribble(
  ~colA, ~colB,
  "a",   1,
  "b",   2,
  "c",   3
)
```

## Tidying Data

Data can be tabulated in one of two ways: A tidy and a messy one.

> If each variable forms a column, each row represents an observation, and each cell a single value, we have **‘tidy data’**. 

Having your data tidied is crucial for facilitating data manipulation, modelling, and visualization!

Here are some guidelines:

1. Don't use values as column headers, all column headers should be variable names.
2. Don't put multiple variables in a single column.
3. Don't store variables in both rows and columns.
4. Avoid storing multiple types of observational units in the same table.

Typically, if the table with your data is wider than long (if it is in the ‘wide format’), it's likely to encounter messy data. For example, `plate_1` is in one (of many possible) messy formats: The concentration is spread along the header row, not in a separate column. 

A tidier ‘wide format’ of the same data would look like that:

```{r echo=FALSE}
plate_1 %>% 
  pivot_longer(cols = -one_of("replicate_id", "sample_id"), 
               names_to  = "concentration", values_to = "intensity") %>% 
  pivot_wider(names_from = sample_id, values_from = intensity) %>% 
  mutate(replicate_id  = str_c("<b>", replicate_id, "</b>"),
         concentration = str_c("<b>", concentration, "</b>"),) %>% 
  head(7) %>% 
  knitr::kable(format = "markdown", align = "c")
```

_(table abridged)_

The longer the table, the tidier the data. Here is the ‘long format’ of `plate_1`:

```{r echo=FALSE}
plate_1 %>% 
  pivot_longer(cols = -one_of("replicate_id", "sample_id"), 
               names_to= "concentration", values_to = "intensity") %>% 
  mutate(sample_id     = str_c("<b>", sample_id, "</b>"),
         replicate_id  = str_c("<b>", replicate_id, "</b>"),
         concentration = str_c("<b>", concentration, "</b>"),) %>% 
  head(9) %>% 
  knitr::kable(format = "markdown", align = "c")
```

_(table abridged)_

The `tidyr` package (part of the `tidyverse`) provides functions to help you tidy messy data.

### Interconverting Wide and Long Table Formats

The degree to which you want to make your table ‘longer’ than ‘wide’ depends on the manipulation you want to perform. Typically, the ‘longest format’ serves as the linchpin to produce the ‘wider formats’.

For your convenience, this is an overview of the different functions that are around to interconvert wide and long table formats. (Including other packages, in case you encounter one of these or need help on related topics.)

| action | `tidyr` ≥ v1.0 | `tidyr` < v1.0 | `data.table` | `reshape`/`reshape2` |
|--------|------------------|------------------|--------------|----------------------|
| make wide table long (‘melting’) | `pivot_longer` | `gather` | `melt`  | `melt` |
| make long table wide (‘casting’) | `pivot_wider`  | `spread` | `dcast` | `acast`, `dcast` |

As many people don't find the other functions intuitive, we will use `tidyr::pivot_longer` and `tidyr::pivot_wider`. 

```{r}
plate_1 %>% 
  # make wide table long
  pivot_longer(cols = conc_1:conc_0,        # selection of columns
               names_to  = "concentration", # name that describes the parameter in the selected headers
               values_to = "intensity"      # name that describes the value in the selected cells
  ) %>% 
  # print only top rows
  head(3)
```

Instead of selecting the columns `conc_1:conc_0`, we could have specified the columns which _not_ to gather; these are somtimes called the ‘identifying (id) columns’. The following commands evaluate to the same result.

```{r eval=FALSE}
plate_1 %>% 
  # select explicitly negative
  pivot_longer(cols = -c("sample_id", "replicate_id"))

plate_1 %>% 
  # select by column index
  pivot_longer(cols = 3:7)

plate_1 %>% 
  # select by name range
  pivot_longer(cols = conc_1:conc_0)

plate_1 %>% 
  # select explicitly positive
  pivot_longer(cols = c("conc_1", "conc_2", "conc_3", "conc_4", "conc_0"))

plate_1 %>% 
  # select using pattern
  pivot_longer(cols = starts_with("conc"))
```

We discuss selecting columns by name later.

For now, let's continue with widening/spreading/casting the long/gathered/molten representation of `plate_1`! Maybe, we would like to see the measured intensities for `control` and `treatment_A` side-by-side given each concentration and the replicate.

```{r}
plate_1 %>% 
  # make wide table long
  pivot_longer(conc_1:conc_0, names_to = "concentration", values_to = "intensity") %>% 
  # make long table wide
  pivot_wider(names_from  = sample_id, # column to use content as new column headers
              values_from = intensity  # column to use content to populate the new cells with
             ) %>% 
  # print only top rows
  head(3) 
```

This format of `plate_1` could be useful if we wanted to subtract the control measurements from the treated samples. Here, `dplyr::mutate(...)` creates a new column or modifies an exisiting column using the specified operations. (More on this later, too.)

```{r}
plate_1 %>% 
  # make wide table long
  pivot_longer(conc_1:conc_0, names_to = "concentration", values_to = "intensity") %>% 
  # make long table wide
  pivot_wider(names_from = sample_id, values_from = intensity) -> plate_1.wide

plate_1.wide %>% 
  # make new column with corrected treatment_A values
  mutate(tratment_A.corrected = treatment_A - control) %>% head(3)

plate_1.wide %>% 
  # operate in place; not always advisable
  mutate(treatment_A = treatment_A - control) %>% head(3)
```

You can check out more examples in the `tidyr` [vignette on pivoting](https://tidyr.tidyverse.org/dev/articles/pivot.html).

### Uniting and Splitting Columns

`tidyr::unite(...)` takes the values form multiple columns and pastes them together side-by-side.

```{r}
plate_1 %>% 
  unite(col = "experiment",      # column to create
        sample_id, replicate_id, # columns to paste, can be unquoted
        sep = "/"                # separator, "_" is the default
  ) %>% 
  # print only top rows
  head(3)
```

The `tidyr::separate(column)` function does the reverse. By default, any non-alphanumeric character will be used to split the column, which can result in missing pieces to be replaced with `NA`.

```{r}
plate_1 %>% 
  separate(col  = sample_id,                 # column to split, can be unquoted
           into = c("treatment", "specimen") # names of the new columns
  ) %>% 
  # print only top rows
  head(3)

plate_1 %>% 
  unite("experiment", sample_id, replicate_id, sep = "/") %>% 
  separate(experiment, into = str_c("piece_", 1:5)) %>% 
  # print only top rows
  head(3)
```

## Hands-On Exercise

The data for another plate has been saved in a file called ‘plates.RData’. When you load this file, you should see a data set called `plate_4` and a named vector called `dose` in the ‘Global Environment’.

```{r}
load(file = "./part_10-working_with_tables_files/plates.RData")
```

1. Combine `plate_4` with `plate_data`.

```{r include=FALSE}
plate_4 %>% 
  # make "sample_id" and "concentration" from "exp_id"
  separate(exp_id, into = c("sample_id", "concentration"), sep = "\\.") %>% 
  # make long table in order to be able to make wide table with concentrations 
  # spread along columns
  pivot_longer(starts_with("rep"), names_to = "replicate_id") %>% 
  # make wide
  pivot_wider(id_cols = ends_with("id"), names_from = concentration, values_from = value) %>% 
  # combine
  bind_rows(plate_data) %>% 
  # re-order and save
  arrange(sample_id, replicate_id) -> plate_data
```

2.  Tidy `plate_data` and replace the concentration with actual values.

```{r include=FALSE}
plate_data %>% 
  pivot_longer(starts_with("conc"), names_to = "concentration", values_to = "intensity") %>% 
  mutate(concentration = dose[concentration]) %>% 
  # re-order and save
  arrange(sample_id, replicate_id, concentration) -> plate_data
```

```{r include=FALSE}
saveRDS(plate_data, file = "part_11-plate_data.rds")
 ```